manuscript_outline

### 2025-06-12 docs/manuscript/paper.Rmd

Historically, models of developmental phenology in wild populations have relied on the accumulation of degree days (thermal sums) above a species-specific threshold. While widely used, degree-day models are fundamentally linear and fail to account for the non-linear nature of development, leading to biased predictions under fluctuating temperature regimes [@neuheimer2009]. This is particularly problematic for long-developing species or for organisms in variable climates, where embryos incubating at colder temperatures often require fewer thermal units to complete development than those incubating in warmer environments. Moreover, traditional models often assume temporally invariant developmental rates and do not incorporate diel temperature variability, which can significantly affect development in systems with large day–night thermal swings. Finally, developmental reaction norms may vary across populations due to local adaptation or phenotypic plasticity, further complicating predictions [@sparks2017; @angilletta2009].

@@

Angilletta, M. J. (2006). Estimating and comparing thermal performance curves. Journal of Thermal Biology.

Angilletta, M. J. (2009). Thermal Adaptation: A Theoretical and Empirical Synthesis.

Schoolfield, R. M., Sharpe, P. J. H., & Magnuson, C. E. (1981). Non-linear regression of biological temperature-dependent rate models. Journal of Theoretical Biology.

Neuheimer, A. B., & Taggart, C. T. (2007). The growing degree-day and fish size-at-age: the overlooked metric. Canadian Journal of Fisheries and Aquatic Sciences.

@@

where $$i$$ is the index for each day, $$a$$ and $$b$$ are empirically derived parameters, and temperature is the average for that day. Development is complete once the cumulative sum of daily effective values reaches one: $$\sum_{i =1}^nEffectiveValue_i = 1$$

@@

Our results indicate developmental rates were highly variable over different physical habitats (streams originating in lowland hills compared with those originating in mountain ranges with permanent snow fields), with sites in Olympic National Park consistently having the longest development times and the greatest interannual variability (SD ≈ 11 days), while lowland streams in the Willapa Hills developed fastest with minimal variability (SD < 3 days). These results demonstrate the capacity of **hatchR** to detect spatiotemporal phenological variation and forecast potential shifts under shifting climate patterns.

@@

As with any model of developmental timing, it is important to recognize that temperature is not the sole determinant of phenological events. Other environmental cues—such as photoperiod, substrate quality, hydrology, salinity, or exposure to contaminants (e.g., metals or endocrine-disrupting chemicals)—can modulate developmental rates or induce thresholds not captured in thermal response curves (NEED REF). Biotic cues may also play a role; for example, the presence of predators can delay hatching in some amphibians as an adaptive strategy [e.g., @sih2001]. While **hatchR** focuses specifically on temperature-dependent development, it can be integrated with complementary datasets or experimental manipulations to investigate these more complex, context-dependent interactions.

Sih, A., Bell, A. M., & Kerby, J. L. (2004). Two stressors are far deadlier than one. Trends in Ecology & Evolution, 19(6), 274–276.

>>>>

Historically, models of developmental phenology in wild populations have relied on the accumulation of degree days (thermal sums) above a species-specific threshold. While widely used, degree-day models are fundamentally linear and fail to account for the non-linear nature of development, leading to biased predictions under fluctuating temperature regimes [@neuheimer2007]. This is particularly problematic for long-developing species or for organisms in variable climates, where embryos incubating at colder temperatures often require fewer thermal units to complete development than those incubating in warmer environments. Moreover, traditional models often assume temporally invariant developmental rates and do not incorporate diel temperature variability, which can significantly affect development in systems with large day–night thermal swings. Finally, developmental reaction norms may vary across populations due to local adaptation or phenotypic plasticity, further complicating predictions [@sparks2017; @angilletta2009].

@@

@@

where *i* is the index for each day, *a* and *b* are empirically derived parameters, and temperature is the average for that day. Development is complete once the cumulative sum of daily effective values reaches one: $$\sum_{i =1}^nEffectiveValue_i = 1$$

@@

Our results indicate developmental rates were highly variable over different physical habitats (streams originating in lowland hills compared with those originating in mountain ranges with permanent snow fields), with sites in Olympic National Park consistently having the longest development times and the greatest interannual variability (SD ~ 11 days), while lowland streams in the Willapa Hills developed fastest with minimal variability (SD < 3 days). These results demonstrate the capacity of **hatchR** to detect spatiotemporal phenological variation and forecast potential shifts under shifting climate patterns.

@@

As with any model of developmental timing, it is important to recognize that temperature is not the sole determinant of phenological events. Other environmental cues—such as photoperiod, substrate quality, hydrology, salinity, or exposure to contaminants (e.g., metals or endocrine-disrupting chemicals)—can modulate developmental rates or induce thresholds not captured in thermal response curves (NEED REF). Biotic cues may also play a role; for example, the presence of predators can delay hatching in some amphibians as an adaptive strategy [e.g., @sih2004]. While **hatchR** focuses specifically on temperature-dependent development, it can be integrated with complementary datasets or experimental manipulations to investigate these more complex, context-dependent interactions.
